Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2020-05-13 Nicolas.tex
YC add

\author{
    N. Nicolas \thanks{n.nicolas@ipnl.in2p3.fr, equal contribution} \inst{1} 
    \and M. Rigault \thanks{m.rigault@ipnl.in2p3.fr, equal contribution} \inst{2}
    \orcid{0000-0002-8121-2560}
    \and\\ Y. Copin \inst{1}
    \and R. Graziani \inst{2}
    %HERE ALPHABETIC ORDER
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and J. Nordin\inst{4}
    \and A.~G. Kim\inst{3}
    \and Y.-L. Kim\inst{1}
    \and Saul Perlmutter\inst{3}
    % Gerard ne souhaite pas etre co-auteur.
}

@@

\abstract{The true nature of type Ia supernovae (SNe~Ia) remains largely
    unknown, and as survey statistics increase, the question of astrophysical
    systematic uncertainties rises, notably that of the SN~Ia population
    redshift drift. In this paper, we study the redshift evolution of SN~Ia
    \textsc{\texttt{SALT2.4}} lightcurve stretch, a purely intrinsic SN
    property, to probe this drift. The SN stretch has been shown to strongly
    correlate with the SN's environment, notably with stellar age tracers. We
    model the underlying stretch distribution as a function of reshift, using
    the evolution of fraction of young and old SNe~Ia as predicted by
    \cite{rigault2018}, and assuming constant underlying stretch distribution
    for each age population made of Gaussian-mixtures.  We test our prediction
    against data from the literature chosen to have negligible selection effects
    to ensure that any observed drift is indeed astrophysical and not
    observational. We clearly demonstrate that the underlying SNe~Ia stretch
    distribution is evolving as a function of redshift, and that the young/old
    drifting model is a much better description of the data than any
    non-drifting model, including the sample-based asymmetric distributions
    usually used to correct Malmquist bias. Our favored underlying stretch model
    is bimodal: a high-stretch mode shared by both young and old environments,
    and a low-stretch mode exclusive to old environments. The precise impact of
    the redshift evolution of the underlying SNe Ia population on cosmology
    remains to be studied. Yet, the astrophysical drift of the SN stretch
    distribution does affect current Malmquist bias corrections and thereby
    distances derived from SN affected by selection effects. We highlight that
    such a bias will increase with surveys covering larger redshift ranges,
    which is particularly important for LSST.}

@@

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2018} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). The spectroscopic follow-up was done over a redshift
range of $0.02 < z < 0.09$~\citep[as in][]{rigault2018}, while the search was
much deeper. \mr{These 114 SNfactory SNe~Ia are thus part of the volume limited
part of the survey (Aldering in prep.).} As such, these SNe~Ia are assumed to be
a random sampling of the underlying SN population. The SNfactory sample is
particularly useful for studying SN property drift, as it enables us to have a
large SN~Ia sample at $z<0.1$.

@@

\mr{Following \cite{rigault2018}, we set $p_y^i$ as one minus the cumulative
distribution function of the $\log(\mathrm{LsSFR})$ at
$-10.82\,\mathrm{dex}$. This is defining the probability that a SN projenitor
is classified as young (see further discussion about this assumption in Briday
et al. in~prep). Since the drift is negligible within the SNfactory dataset
($\Delta\delta(z)<3\%$ from the highest to the lowest redshift of the dataset),
one can readily use $y^i = p_y^i$ to minimize Eq.~\ref{eq:likelihood} with
respect to $\vec{\theta}$. Results on fitting the SNfactory SNe with this model
are shown Table~\ref{tab:modelresults} and illustrated in
Fig.~\ref{fig:stretchlssfr}.}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.7\linewidth]{Article_figures/stretchevol_all_vs_snf-hist_on.pdf}
    \caption{Evolution of the mean SN \textsc{\texttt{SALT2.4}} stretch ($x_1$)
        as a function of redshift. Markers show the mean stretch measured in
        redshift bins of equal sample size, indicated in light gray at the
        bottom of each redshift bins. Full and light markers are used when
        considering the fiducial or the conservative samples, respectively. The
        orange horizontal line represents the mean stretch of the fiducial
        sample, illustrating the expectation if the SN stretch distribution is
        not evolving with redshift. Best fits of our Base drifting model are
        shown as blue, dashed-blue and gray, when fitted on the fiducial sample,
        the conservative one or the SNfactory dataset only, respectively; all
        are compatible. The light-blue band illustrates the amplitude of the
        error (incl. covariance) of the best fit model when considering the
        fiducial dataset. \mr{The redshift distributions of each sample are
        shown as light-histograms, see lower left legend.}}
    \label{fig:modelall}
\end{figure*}

@@

Finally, for the sake of completeness, we also consider Gaussian
and Asymmetric Gaussian non-drifting models.

@@

Since the various models have different degrees of freedom, we use the Akaike
Information Criterion \citep[AIC, e.g.][]{burnham2004} to compare their ability
to properly describe the observations. This estimator penalizes extra degrees of
freedom to avoid over-fitting the data, and is defined as follow:
\begin{equation}
    \mathrm{AIC} = -2\ln(L) + 2k
\end{equation}
where $-2\ln(L)$ is derived by minimizing Eq.~\eqref{eq:likelihood}, and $k$ is
the number of free parameters to be adjusted. The reference model is the one
with the smallest AIC; in comparison to this model, the models with
$\Delta\mathrm{AIC}<5$ are coined acceptable, the ones with
$5<\Delta\mathrm{AIC}<20$ are unfavored, and those with $\Delta\mathrm{AIC}>20$
are deemed excluded. This roughly corresponds to 2, 3 and 5~$\sigma$ limits for
a Gaussian probability distribution. \mr{However, we highlight that the best of
the compared models might not be the true underlying model.}

@@

Surprisingly, the sample-based Gaussian asymmetric modeling used by current
implementations of the BBC technique \citep{scolnic2016, kessler2017} has one of
the highest AIC value in our analysis (see Section~\ref{sec:results}). While its
$-2\ln(L)$ is the smallest of all redshift-independent models (but still $-11.5$
worse than the reference Base model), it is strongly penalized for requiring
15~free parameters ($\mu_0, \sigma_{\pm}$~for each of the 5~samples of the
analysis). Hence, its $\Delta\mathrm{AIC}<-20$, which could be interpreted as a
probability $p=2\times 10^{-7}$ of being an as good representation of the data
as the Base model.
\mr{Remark that the Bayesian Information Criterion ($\mathrm{BIC} = -2\ln(L) +
k\ln(n)$, with $n$ is the number of datapoint) might be better suited than AIC
when comparin models to the sample-based one since it explicitely accounts for
the fact that each sample is fitted separately; the sample-based model BIC is
the sum of the BIC per sample. Doing so we find $\Delta\mathrm{BIC}=-48$, also
excluding the sample-base asymmetric Gaussian model as an as good description of
the data then the Base-model.}

@@

To the best of our knowledge, a SN~Ia stretch redshift drift modeling has never
been explicitly used in cosmological analyses, though bayesian hierarchy
formalism such as UNITY \citep{rubin2015}, BAHAMAS \citep{shariff2016} or Steve
\citep{hinton2019} can easily allow it; see e.g., section 1.3 and 2.5 of
\cite{rubin2015}. Not doing so is a second order issue for SN cosmology, as it
only affects the way one accounts for Malmquist bias. Indeed, as long as
Phillips' relation \citep{phillips1993} standardization parameter $\alpha$ is
not redshift dependent (a study behind the scope of this paper, but see e.g.
\citealt{scolnic2018a}), the stretch-corrected SNe~Ia magnitudes used for
cosmology are blind to the underlying stretch distribution for complete samples.
However, surveys usually do have significant Malmquist bias at least for the
upper half of their SN redshift distribution. As a consequence, an ill-modeling
of the underlying stretch distribution will bias the SN magnitude derived from
such surveys.

@@

The amplitude of this magnitude bias for cosmology is beyond the scope of this
paper given the complexity of the BBC analysis. It would require a full study
using our Base model (Eq.~\ref{eq:stretchz}) in place of the sample-based
asymmetric modeling as part of the BBC simulations.
\mr{However, we already highlight that even if a non-drifting sample-based model
could provide comparable result in the volume-limited part of the various
samples, these models would differ when extrapolating at higher redshift,
precisely where the  underlying will matter for correcting Malmquist biases.}

@@

In addition to this ``base'' modeling, we have studied various distributions,
including redshift independent models; we also studied the prediction from the
``Beams with Bias Correction'', a Malmquist bias correction algorithm used by
several recent SN~Ia cosmological analyses, implemented as a per-sample
asymmetric Gaussian distribution.

Our conclusions are the following:
\begin{enumerate}

    \item The underlying SN~Ia stretch distribution is redshift dependent, as
        previously suggested by e.g.~\cite{howell2007}. This result is largely
        independent of details on each age-population modeling.

    \item Non-drifting models are indeed excluded as suitable descriptions of
        the data relative to our Base model. This model assumes that: (1) the
        younger population has a unimodal Gaussian stretch distribution, while
        the older population stretch distribution is bimodal, one mode being the
        same as the young one; (2) the evolution of the relative fraction of
        younger and older SNe~Ia follows the prediction made in
        \cite{rigault2018}. This second result strongly supports the existence
        of both young and old SN~Ia populations, in agreement with rate studies.

    \item Using survey-based asymmetric Gaussian distributions, as  done, e.g.,
        in the current implementation of the BBC, is excluded to be good
        descriptions of the data relative to our drifting model. Hence, the
        sample-based approach does not accurately account for redshift drift and
        even less so as survey span larger redshift ranges. We warn that, even
        if extra degrees of freedom might be acceptable given the large
        number of SNe~Ia in cosmological studies, extrapolating the SN property
        distributions from the volume-limited part of a survey to the
        magnitude-limited one affected by sample selection would still be
        inaccurate since the distribution would drift.

@@

    \item Given the current dataset, we suggest the use of the following stretch
        population model as a function of redshift:
        \begin{align}\label{eqconclusion:stretchz}
            X_1\left(z \right) =
             &\,\delta(z)\times\mathcal{N}(\mu_1,\sigma_1^2)\,+\nonumber\\
            (1-&\,\delta(z)) \times  \left[a\times\mathcal{N}(\mu_1,\sigma_1^2) +
            (1-a)\times\mathcal{N}(\mu_2,\sigma_2^2)\right]
        \end{align}
        with: $a=0.51$, $\mu_1=0.37$, $\mu_2=-1.22$, $\sigma_1=0.61$,
        $\sigma_2=0.56$ (see Table~\ref{tab:modelresults}) and using the
        age-population drift modelization $\delta(z)$ with $K=0.87$:
        \begin{align}
            \delta(z) & = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}.
        \end{align}
\end{enumerate}

In this paper, we considered a simple Gaussian mixture modeling, but additional
data free from significant Malmquist bias would enable us to refine it. We note
that samples at the low- and high-redshift ends of the Hubble diagram would be
particularly helpful for this drifting analysis; fortunately this will soon be
provided by the Zwicky Transient Facility \citep{bellm2019, graham2019} and
Subaru and SeeChange SNe~Ia programs, respectively.

>>>>

\author{
    N. Nicolas \thanks{n.nicolas@ipnl.in2p3.fr, equal contribution} \inst{1} 
    \and \mr{M. Rigault} \thanks{m.rigault@ipnl.in2p3.fr, equal contribution} \inst{2}
    \orcid{0000-0002-8121-2560}
    \and\\ \yc{Y. Copin} \inst{1}
    \orcid{0000-0002-5317-7518}
    \and R. Graziani \inst{2}
    %HERE ALPHABETIC ORDER
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and J. Nordin\inst{4}
    \and A.~G. Kim\inst{3}
    \and Y.-L. Kim\inst{1}
    \and Saul Perlmutter\inst{3}
    % Gerard ne souhaite pas etre co-auteur.
}

@@

\abstract{The true nature of type Ia supernovae (SNe~Ia) remains largely
    unknown, and as survey statistics increase, the question of astrophysical
    systematic uncertainties rises, notably that of the SN~Ia population
    redshift drift. In this paper, we study the redshift evolution of SN~Ia
    \textsc{\texttt{SALT2.4}} lightcurve stretch, a purely intrinsic SN
    property, to probe this drift. \yc{\emph{isn't a 'drift' \textbf{exactly} a redshift evolution?}} The SN stretch has been shown to strongly
    correlate with the SN environment, notably with stellar age tracers. We
    model the underlying stretch distribution as a function of redshift, using
    the evolution of fraction of young and old SNe~Ia as predicted by
    \cite{rigault2018}, and assuming constant underlying stretch distribution
    for each age population made of Gaussian mixtures.  We test our prediction
    against \yc{published samples} chosen to have negligible \yc{magnitude} selection effects,
    \yc{so} that any observed \yc{change} is indeed \yc{of astrophysical and not
    observational origin}. We clearly demonstrate that the underlying SN~Ia stretch
    distribution is evolving as a function of redshift, and that the young/old
    drifting model is a much better description of the data than any
    \yc{\sout{non-drifting} constant} model, including the sample-based asymmetric distributions
    usually used to correct Malmquist bias. \yc{The} favored underlying stretch model
    is \yc{the bimodal one introduced in \cite{rigault2018}}: a high-stretch mode shared by both young and old environments,
    and a low-stretch mode exclusive to old environments. The precise impact of
    the redshift evolution of the SN Ia population \yc{intrinsic properties} on cosmology
    remains to be studied. Yet, the astrophysical drift of the SN stretch
    distribution does affect current Malmquist bias corrections and thereby
    distances derived from SN affected by selection effects. We highlight that
    such a bias will increase with surveys covering \yc{increasingly} larger redshift ranges,
    which is particularly important for LSST.}

@@

\yc{\emph{IMHO too many occurrences of `drift' (53!); this could be replaced by '[redshift] evolution' to be opposed to 'constant'.}}

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2018} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). The spectroscopic follow-up was done over a redshift
range of $0.02 < z < 0.09$~\citep[as in][]{rigault2018}, while the search was
much deeper. \mr{These 114 SNfactory SNe~Ia are thus \yc{in} the volume limited
part of the survey (Aldering \yc{et al.}, in prep.)}, \yc{and are therefore} assumed to be
a random sampling of the underlying SN population. The SNfactory sample is
particularly useful for studying SN property drift, as it enables us to have a
large SN~Ia sample at $z<0.1$.

@@

\mr{Following \cite{rigault2018}, we \yc{define?} $p_y^i$ as one minus the cumulative
distribution function of the $\log(\mathrm{LsSFR})$ at
$-10.82\,\mathrm{dex}$. This is defining the probability that a SN progenitor
is classified as young (see further discussion about this assumption \yc{\emph{which assumption, if it's purely your own definition of young vs. old?}} in Briday
et al. in~prep). Since the drift is negligible within the SNfactory dataset
($\Delta\delta(z)<3\%$ from the highest to the lowest redshift of the dataset),
one can readily use $y^i = p_y^i$ to minimize Eq.~\ref{eq:likelihood} with
respect to $\vec{\theta}$. Results on fitting the SNfactory SNe with this model
are shown Table~\ref{tab:modelresults} and illustrated in
Fig.~\ref{fig:stretchlssfr}.} \yc{\emph{Now I'm confused why one does not have \textbf{exactly} $y^i = p_y^i$ when LsSFR measurement is available, but you seem to say it's an assumption.}}

@@

\begin{figure*}
    \centering
    \includegraphics[width=0.7\linewidth]{Article_figures/stretchevol_all_vs_snf-hist_on.pdf}
    \caption{Evolution of the mean SN \textsc{\texttt{SALT2.4}} stretch ($x_1$)
        as a function of redshift. Markers show the mean stretch measured in
        redshift bins of equal sample size, indicated in light gray at the
        bottom of each redshift bins. Full and light markers are used when
        considering the fiducial or the conservative samples, respectively. The
        orange horizontal line represents the mean stretch of the fiducial
        sample, illustrating the expectation if the SN stretch distribution is
        not evolving with redshift. Best fits of our Base drifting model are
        shown as blue, dashed-blue and gray, when fitted on the fiducial sample,
        the conservative one or the SNfactory dataset only, respectively; all
        are compatible. The light-blue band illustrates the amplitude of the
        error (incl. covariance) of the best fit model when considering the
        fiducial dataset. \mr{The redshift distributions of each sample are
        shown as light-histograms, see lower left legend.} \yc{\emph{This is upper panel of Fig.3? Why is it again split in subsamples, while it was not deemed necessary in dedicated Fig.3?}}}
    \label{fig:modelall}
\end{figure*}

@@

Finally, for the sake of completeness, we also consider \yc{constant} Gaussian
and Asymmetric Gaussian \yc{\sout{non-drifting}} models.

@@

Since the various models have different degrees of freedom, we use the Akaike
Information Criterion \citep[AIC, e.g.][]{burnham2004} to compare their ability
to properly describe the observations. This estimator penalizes extra degrees of
freedom to avoid over-fitting the data, and is defined as follow:
\begin{equation}
    \mathrm{AIC} = -2\ln(L) + 2k
\end{equation}
where $-2\ln(L)$ is derived by minimizing Eq.~\eqref{eq:likelihood}, and $k$ is
the number of free parameters to be adjusted. The reference model is the one
with the smallest AIC; in comparison to this model, the models with
$\Delta\mathrm{AIC}<5$ are coined acceptable, the ones with
$5<\Delta\mathrm{AIC}<20$ are unfavored, and those with $\Delta\mathrm{AIC}>20$
are deemed excluded. This roughly corresponds to 2, 3 and 5~$\sigma$ limits for
a Gaussian probability distribution. \mr{However, we highlight that the best of
the compared models might not be the true underlying model.} \yc{\emph{why this sentence? Is there any \textbf{true} intrinsic model?}}

@@

Surprisingly, the sample-based Gaussian asymmetric modeling used by current
implementations of the BBC technique \citep{scolnic2016, kessler2017} has one of
the highest AIC value in our analysis (see Section~\ref{sec:results}). While its
$-2\ln(L)$ is the smallest of all redshift-independent models (but still $-11.5$
worse than the reference Base model), it is strongly penalized for requiring
15~free parameters ($\mu_0, \sigma_{\pm}$~for each of the 5~samples of the
analysis). Hence, its $\Delta\mathrm{AIC}<-20$, which could be interpreted as a
probability $p=2\times 10^{-7}$ of being an as good representation of the data
as the Base model.

\yc{Remark that, when comparing models adjusted on individual subsamples rather than globally, the Bayesian Information Criterion ($\mathrm{BIC} = -2\ln(L) +
k\ln(n)$, with $n$ the number of data points) might be better suited than AIC, since it explicitly accounts for
the fact that each subsample is fitted separately: the sample-based model BIC is
rightfully the sum of the BIC for each sample. Doing so, we find $\Delta\mathrm{BIC}=-48$, 
again \sout{excluding} \emph{refuting} the sample-based asymmetric Gaussian model as being as pertinent as the Base model.}

@@

To the best of our knowledge, a SN~Ia stretch redshift drift modeling has never
been explicitly used in cosmological analyses, though Bayesian hierarchy
formalism such as UNITY \citep{rubin2015}, BAHAMAS \citep{shariff2016} or Steve
\citep{hinton2019} can easily allow it; see e.g., section 1.3 and 2.5 of
\cite{rubin2015}. Not doing so is a second order issue for SN cosmology, as it
only affects the way one accounts for Malmquist bias. Indeed, as long as
Phillips' relation \citep{phillips1993} standardization parameter $\alpha$ is
not redshift dependent (a study behind the scope of this paper, but see e.g.
\citealt{scolnic2018a}), the stretch-corrected SNe~Ia magnitudes used for
cosmology are blind to the underlying stretch distribution for complete samples.
However, surveys usually do have significant Malmquist bias at least for the
upper half of their SN redshift distribution. As a consequence, an ill-modeling
of the underlying stretch distribution will bias the SN magnitude derived from
such surveys.

@@

The amplitude of this magnitude bias for cosmology is beyond the scope of this
paper given the complexity of the BBC analysis. It would require a full study
using our Base model (Eq.~\ref{eq:stretchz}) in place of the sample-based
asymmetric modeling as part of the BBC simulations.
\mr{However, we already highlight that even if a non-drifting sample-based model
could provide comparable result in the volume-limited part of the various
samples, these models would differ when extrapolating at higher redshifts,
precisely where the  underlying \yc{\emph{missing word?}} will matter for correcting Malmquist biases.}

@@

In addition to this ``base'' modeling, we have studied various distributions,
including redshift independent models; we also studied the prediction from the
``Beams with Bias Correction'', a Malmquist bias correction algorithm used by
several recent SN~Ia cosmological analyses, implemented as a per-sample
asymmetric Gaussian distribution. \yc{\emph{of what?}}

Our conclusions are the following:
\begin{enumerate}
    \item The underlying SN~Ia stretch distribution is \yc{significantly} redshift dependent, as
        previously suggested by e.g.~\cite{howell2007}. This result is largely
        independent of details on each age-population modeling.

    \item \yc{\sout{Non-drifting} Redshift-independent} models are indeed excluded as suitable descriptions of
        the data relative to our Base model. This model assumes that: (1) the
        younger population has a unimodal Gaussian stretch distribution, while
        the older population stretch distribution is bimodal, one mode being the
        same as the young one; (2) the evolution of the relative fraction of
        younger and older SNe~Ia follows the prediction made in
        \cite{rigault2018}. This second result strongly supports the existence
        of both young and old SN~Ia populations, in agreement with rate studies. \yc{\emph{ref???}}

    \item Using survey-based asymmetric Gaussian distributions, as done, e.g.,
        in the current implementation of the BBC, is excluded to be good
        descriptions \yc{\emph{rephrase: "using is excluded"?}} of the data relative to our drifting model. Hence, the
        sample-based approach does not accurately account for redshift drift and
        even less so as survey span \yc{increasingly} larger redshift ranges. We \yc{\sout{warn} stipulate} that, even
        if extra degrees of freedom might be acceptable given the large
        number of SNe~Ia in cosmological studies, extrapolating the SN property
        distributions from the volume-limited part of a survey to its
        \yc{Malmquist-biased} magnitude-limited one \yc{\sout{affected by sample selection}} would still be
        inaccurate \yc{\sout{since the distribution would drift} because of the redshift evolution}.

@@

    \item Given the current dataset, we suggest the use of the following stretch
        population model as a function of redshift:
        \begin{align*}
        \label{eqconclusion:stretchz}
            X_1\left(z \right) =
             &\,\delta(z)\times\mathcal{N}(\mu_1,\sigma_1^2)\,+\nonumber\\
            (1-&\,\delta(z)) \times  \left[a\times\mathcal{N}(\mu_1,\sigma_1^2) +
            (1-a)\times\mathcal{N}(\mu_2,\sigma_2^2)\right]
            \tag{\ref{eq:stretchz}}
        \end{align*}
        with $a=0.51$, $\mu_1=0.37$, $\mu_2=-1.22$, $\sigma_1=0.61$,
        $\sigma_2=0.56$ (see Table~\ref{tab:modelresults}), and using the
        age-population drift model         \begin{align*}
            \delta(z) & = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}
            \tag{\ref{eq:delta}}
        \end{align*}
        with $K=0.87$.
\end{enumerate}

In this paper, we considered a simple Gaussian mixture modeling, but additional
data free from significant Malmquist bias would enable us to refine it \yc{as necessary}. We note
that samples at the low- and high-redshift ends of the Hubble diagram would be
particularly helpful for this drifting analysis; fortunately this will soon be
provided by the Zwicky Transient Facility \citep{bellm2019, graham2019} and
Subaru and SeeChange SNe~Ia programs, respectively. \yc{\emph{cannot say who's who in 'A and B and C'.}}
